---
title: "Elementary Statistics with R"
author: 
date: 
output:
   html_document:
     fig_width: 8
     fig_height: 8
     keep_md: true
     toc: true
     toc_depth: 2
---
**Personal notes. Reference: r-tutor**
http://www.r-tutor.com/elementary-statistics

```{r message=FALSE}
#library(MASS)
library(dplyr)
library(xtable)

print_table <- function(df,n=6){
  print(xtable::xtable(head(df,n=n)), type='html')
}
```

# Categorical Data
http://www.r-tutor.com/elementary-statistics/qualitative-data
```{r}
#https://blog.rstudio.org/2016/03/24/tibble-1-0-0/
painters <- dplyr::as_data_frame(MASS::painters)
dim(painters)
head(painters)
```


```{r results='asis'}
print_table(painters)
```

## Get frequency counts using table
```{r}
school = painters$School    
school
as.numeric(school)

# apply table function to get counts
school.freq = table(school)
school.freq

# use cbind to print vertically
cbind(school.freq)

# convert to relative frequency distribution
default_options = options()
school.relfreq <- school.freq / nrow(painters) 
cbind(school.relfreq)

# change digit output
options(digits=1)
school.relfreq

# reset options
options(default_options)
```

## Plot as bar/pie charts

```{r fig.height=4}
par(mfrow=c(1,3))
barplot(school.freq)
barplot(school.relfreq)
colors = c("red", "yellow", "green", "violet", "orange", "blue", "pink", "cyan") 
barplot(school.freq,col=colors)

# pie char
par(mfrow=c(1,2))
pie(school.freq)              
pie(school.freq,col=colors) 
```

# Quantitative Data
http://www.r-tutor.com/elementary-statistics/quantitative-data
```{r}
#https://blog.rstudio.org/2016/03/24/tibble-1-0-0/
faithful <- dplyr::as_data_frame(faithful)
dim(faithful)
head(faithful)
```


```{r results='asis'}
print_table(faithful)
```

```{r}
duration <- faithful$eruptions
range(duration) # min/max
breaks = seq(1.5, 5.5, by=0.5)    # half-integer sequence 
breaks
```

## Discretize with cut
- cut divides the range of x into intervals and codes the values in x according to which interval they fall. 
- The leftmost interval corresponds to level one, the next leftmost to level two and so on.
```{r}
duration.cut = cut(duration, breaks, right=FALSE) 
duration.cut[1:5]

# now we can apply the table technique to get frequency counts
duration.freq = table(duration.cut) 
duration.relfreq = table(duration.cut) / length(duration.cut)

# add cdf
duration.cumfreq = cumsum(duration.freq)
duration.cumrelfreq = duration.cumfreq / nrow(faithful) 


cbind(duration.freq, duration.relfreq,duration.cumfreq,duration.cumrelfreq) 
duration_df <- data.frame(cbind(duration.freq, 
                                duration.relfreq,
                                duration.cumfreq,
                                duration.cumrelfreq ) )
duration_df
colnames <- names(duration_df) <- c('freq','relfreq','cumfreq','duration.cumrelfreq')
duration_df
# convert to dataframe
# duration_df <- data.frame( freq = as.vector(duration.freq),
#                            relfreq = as.vector(duration.relfreq),
#                            row.names = row.names(duration.freq))
# duration_df <- as_data_frame(duration_df)
# duration_df
```

## Plotting 
```{r fig.height=4}
par(mfrow=c(1,3))
hist(duration)
hist(duration,right=FALSE) # intervals closed on the left

# more styles
hist(duration,
     col=colors,
     main="Old Faithful eruptions", # title text
     xlab = "Duration minutes") # xaxis label

par(mfrow=c(1,2))
cumfreq0 = c(0, cumsum(duration.freq)) 
cumfreq0
plot(breaks, cumfreq0, main = "Old Faithful eruptions", 
     xlab="duration minutes",ylab = "cumulative eruptions")
lines(breaks,cumfreq0) # connect the dots

# alternatively, create an interpolation funcion Fn using ecdf
# ecdf: Compute an empirical cumulative distribution function
Fn = ecdf(duration) 
plot(Fn, main="Old Faithful Eruptions", xlab="Duration minutes", 
    ylab="Cumulative eruption proportion") 

# stem and leaf plot
stem(duration) 

# scatter plot
plot(faithful$eruptions, faithful$waiting,            # plot the variables 
      xlab="Eruption duration",        # x−axis label 
      ylab="Time waited")              # y−axis label 

# enhanced solution
# (abline: Add Straight Lines to a Plot)
plot(faithful$eruptions, faithful$waiting,            # plot the variables 
      xlab="Eruption duration",        # x−axis label 
      ylab="Time waited")              # y−axis label 
# abline(lm( faithful$waiting ~ faithful$eruptions))
abline(lm( waiting ~ eruptions, data=faithful))
```

# Basic statistics
http://www.r-tutor.com/elementary-statistics/numerical-measures

## Scalar statistics
```{r}
head(faithful)

# === mean, median, sd, var,range, iqr ===
mean(duration)
median(duration)
sd(duration)
var(duration)
range(duration) # max - min
IQR(duration) # max - min

# repeat using dataframes
faithful %>% apply(2,mean)
faithful %>% summarise(mean(eruptions),mean(waiting))

# above is tedious if you want to apply same func to all cols....
# but we can use summarise_each
# http://stackoverflow.com/questions/21644848/summarizing-multiple-columns-with-dplyr
# http://stackoverflow.com/questions/21295936/can-dplyr-summarise-over-several-variables-without-listing-each-one
faithful %>% summarise_each(funs(mean,median,sd,var,IQR)) %>% t()
```


## Quantiles, covariance, correlation
```{r}
# ===  quantiles  ===
quantile(duration)
quantiles <- rbind(quantile(faithful$eruptions),quantile(faithful$waiting))
quantiles
row.names(quantiles) <- c('duration','waiting')
quantiles

# data frame approach?
# - use lapply
#http://stackoverflow.com/questions/17020293/quantiles-of-a-data-frame
lapply(faithful, quantile)
as.data.frame(lapply(faithful, quantile))

# === covariance ===
# covariance value
cov(faithful$eruptions,faithful$waiting)

# correlation value
cor(faithful$eruptions,faithful$waiting)

# covariance matrix
cov(faithful)

# correlation matrix
cor(faithful)
```

## Higher order moments, skewness, kurtosis
```{r}
# 3rd order moment
e1071::moment(duration, order=3, center=TRUE) 
e1071::skewness(duration) 
e1071::kurtosis(duration) 
```

# (incomp) Probability Distributions
http://www.r-tutor.com/elementary-statistics/probability-distributions

# (incomp) Interval estimation
http://www.r-tutor.com/elementary-statistics/interval-estimation

# (incomp) Hypothesis testing
http://www.r-tutor.com/elementary-statistics/hypothesis-testing

# (incomp) Type-II Errors
http://www.r-tutor.com/elementary-statistics/type-2-errors

# (incomp) Inference about two populatiosnr
http://www.r-tutor.com/elementary-statistics/inference-about-two-populations

# (incomp) Goodness of Fit
http://www.r-tutor.com/elementary-statistics/goodness-fit


# (incomp) Probability distributions
http://www.r-tutor.com/elementary-statistics/probability-distributions

# (incomp) ANOVA
http://www.r-tutor.com/elementary-statistics/analysis-variance

# (incomp) Nonparametric methods
http://www.r-tutor.com/elementary-statistics/non-parametric-methods

# Simple Linear Regression
http://www.r-tutor.com/elementary-statistics/simple-linear-regression
```{r}
eruption.lm = lm(eruptions ~ waiting, data=faithful) 
eruption.lm
ls.str(summary(eruption.lm))
summary(eruption.lm)


# get coefficients
coeffs = coefficients(eruption.lm)
coeffs
class(coeffs)

# prediction at test point
predict(eruption.lm, data.frame(waiting=40))
predict(eruption.lm, data.frame(waiting=80))

# coefficient of determination
summary(eruption.lm)$r.squared

#Further detail of the summary function for linear regression model
help(summary.lm)
```

## Significance test and confidence interval for linear regression
```{r}
# see coefficients and f-stast
summary(eruption.lm)

# test point and confidence interval (95%)
newdata <- data.frame(waiting=80)

# === prediction vs confidence ===
#http://stats.stackexchange.com/questions/16493/difference-between-confidence-intervals-and-prediction-intervals
#https://www.r-bloggers.com/the-uncertainty-of-predictions/
predict(eruption.lm, newdata, interval="confidence") 
predict(eruption.lm, newdata, interval="prediction") 

# Further detail of the predict function for linear regression model
help(predict.lm)
```

## Residual analysis
$$\text{residual} = y - \hat{y}$$
The **standardized residual** is the residual divided by its standard deviation. 
$$\text{standardized residual } i = \frac{\text{residual } i}{\text{std-dev of resid } i}$$

```{r fig.height=4}
# residual
eruption.res = resid(eruption.lm) 

# plot residual against observed values
par(mfrow=c(1,3))
plot(faithful$waiting, eruption.res, 
     ylab="Residuals", xlab="Waiting Time", 
     main="Old Faithful Eruptions") 
abline(0, 0)                  # the horizon 

# standardized residual
eruption.stdres = rstandard(eruption.lm) 
plot(faithful$waiting, eruption.stdres, 
     ylab="Standardized Residuals", xlab="Waiting Time", 
     main="Old Faithful Eruptions") 
abline(0, 0)                  # the horizon 

# check normality of residual via qqplot
qqnorm(eruption.stdres, 
   ylab="Standardized Residuals", 
   xlab="Normal Scores", 
   main="Old Faithful Eruptions") 
qqline(eruption.stdres) 
```

# (incomp) Multiple linear regression
http://www.r-tutor.com/elementary-statistics/multiple-linear-regression

# Logistic regression
http://www.r-tutor.com/elementary-statistics/logistic-regression

## glm
```{r}
mtcars <- as_data_frame(mtcars)
head(mtcars)

am.glm = glm(formula=am ~ hp + wt, data=mtcars, family=binomial) 

newdata = data.frame(hp=120, wt=2.8) 

help(predict.glm)

# default - scale of linear predictors (prob on logit scale)
predict(am.glm, newdata, type= 'link')    

# scale of response variable (predicted probability)
predict(am.glm, newdata, type='response') 

predict(am.glm, newdata, type='terms') 
```

## Significance test for glm
http://www.r-tutor.com/elementary-statistics/logistic-regression/significance-test-logistic-regression
```{r}
summary(am.glm) 
help("summary.glm")
```


